Relationships in the extremes and their influence on competition and coexistence

### 2021-09-30 paper/Paper.Rnw
spelling fixes in Theory

%Goal para
We previously demonstrated \citep{Ghosh_2020_AER} that ATAs exist and are common in nature, and we argued above 
that it is plausible both that climate change may alter environmental ATAs, and that ATAs may influence species 
coexistence. Thus we seek to explore the consequences of ATAs for competition, coexistence, and the maintenance of 
biodiversity, by means of the following specific goals. 
G1) To augment MCT to show formally how ATAs plays a role in coexistence. 
MCT decomposes a measure of coexistence into components due to each of several mechanisms. 
We address how ATA contributes to these components. 
We here examine mechanisms of coexistence that apply in models exhibiting solely temporal variation; spatial variation can be considered in future work.
G2) To explore, using the simple lottery model, the relative magnitudes of the contributions to coexistence that come
from ATAs compared to other, known mechanisms of coexistence. This is to help illuminate whether ATA contributions to coexistence 
are likely to be negligible or meaningfully large. 
G3) To explore, using our simple model, whether species can coexist with
ATAs but not without, or vice versa. This is to begin to illuminate whether ATA contributions to biodiversity maintenance 
may be negligible or meaningfully large.
G4) To provide an empirical example of species coexistence which highlights the role ATAs can play.
In the Discussion, we revisit the possibility that climate change will alter
ATAs of environmental variables and thereby influence competition and species coexistence.
Overall, our study presents a new mechanism of species coexistence which may be influenced by climate change, 
and a means of understanding its theoretical importance and its importance for real ecosystems.

@@

We here pursue goal G1 from the Introduction: to augment MCT to illuminate how ATA
plays a role in coexistence. For pedgogical clarity, we develop each step of our new theory for a general 
two-species model, and for the lottery model (Introduction).
MCT and its recent computational extensions \citep{Ellner_2016,Ellner_2019} 
quantify the contributions of 
multiple mechanisms to coexistence. But we show below
that, for purely temporal variation, only one mechanism relates to tail associations: storage effects. 
So we begin by defining storage effects for a general two species model and 
for the lottery model. Our mathematical notation parallels
that of \cite{Ellner_2016,Ellner_2019}.

For a general model, it is assumed that the growth rate $r_i(t)=\ln(N_i(t+1)/N_i(t))$ 
of species $i$ can be written as an increasing function of
an environment-dependent factor $E_i(t)$ and as a decreasing function of a quantity representing
competitive pressure on species $i$, $C_i(t)$. As for the lottery model example (Introduction), we here assume, for simpicity, 
that the $E_i(t)$ are iid through time, and likewise for the
$C_i(t)$. Though this assumption can be circumvented, we make the assumption because it simplifies the
presentation of new ideas.

@@

We now specify how to decompose storage effects into
contributions due to ATAs and due
to $EC$ correlation \emph{per se}. 
Storage effects for species $i$ were computed by comparing
differences between invader and resident mean growth rates for the original model and for a 
modified model for which the components of $(E_i,C_{i \bs i})$ and those of $(E_j,C_{j \bs i})$
were rendered independent of each other [see (\ref{eq:storage_effects_gen})]. 
Because these components were rendered 
completely independent in the modified model, storage effects quantify the contribution of the 
whole $EC$ relationship, in all its aspects, to the invader/resident growth rate
difference. But we can instead consider distributions $(E_i^\nat,C_{i \bs i}^\nat)$ and 
$(E_j^\nat,C_{j \bs i}^\nat)$ with the same
marginal distributions, respectively, as $(E_i,C_{i \bs i})$ and $(E_j,C_{j \bs i})$,
and with some but not all aspects of the relationship 
between the components altered relative to the original distributions. Specifically, asymmetries of tail association are removed -- see below for how this is done. We then define 
$\olrn_{i \bs i}$ and $\olrn_{j \bs i}$ based on a modified model using  $(E_i^\nat,C_{i \bs i}^\nat)$
and $(E_j^\nat,C_{j \bs i}^\nat)$, and we consider the decomposition 
\begin{align}
\Delta_i^{(EC)} &= [(\olr_{i\bs i} - q_{ij} \olr_{j \bs i})-(\olrn_{i \bs i}-q_{ij} \olrn_{j \bs i})]+
[(\olrn_{i \bs i}-q_{ij} \olrn_{j \bs i})-(\olrs_{i \bs i}-q_{ij} \olrs_{j \bs i})] \label{eq:storage_effects_further1}\\
&= \Delta_i^{[E \nat C]} + \Delta_i^{[EC]}.\label{eq:storage_effects_further}
\end{align}
The term $\Delta_i^{[E \nat C]}$ is the component of storage effects due to 
ATAs between $E_i$ and $C_{i \bs i}$, and between $E_j$ and $C_{j \bs i}$.
The notation $\nat$ was chosen because this new symbol is part of the symbol $\shp$ (i.e., two of its four lines), 
corresponding to the removal of part of the relationship between $E$ and $C$ components (namely, ATAs).
The new symbol should correspondingly be pronounced ``partial sharp''.
The term $\Delta_i^{[EC]}$ is the component of storage effects due to $EC$ correlation \emph{per se}, remaining after 
having removed the effects of ATAs.
The brackets in the superscripts of $\Delta_i^{[E \nat C]}$ and $\Delta_i^{[EC]}$ are intended to indicate that 
(\ref{eq:storage_effects_further}) is a decompostion within a decomposition, i.e., (\ref{eq:storage_effects_further}) is a decomposition
of storage effects, which are themselves a term within the decomposition provided by MCT \citep{Ellner_2019} -- see 
below for additional details.

How are asymmetries of tail association removed while leaving $E$ and $C$ marginal distributions and overall 
correlation unchanged? We assume, for simplicity, that the cummulative distribution functions (CDFs) of $E_i$ and 
$C_{i \bs i}$, which we denote $F_{E_i}$ and $F_{C_{i \bs i}}$, respectively, are strictly monotonic and continuous. 
Letting $\varphi$ be the CDF of a standard normal distribution, we consider the random variables
$\varphi^{-1} \circ F_{E_i} (E_i)$ and $\varphi^{-1} \circ F_{C_{i \bs i}}(C_{i \bs i})$, which are standard normally
distributed, although they are not necessarily jointly bivariate normal.  Let $\rho_i$ be the covariance of
these random variables. We then define a bivariate normal random variable $(e_i,c_{i \bs i})$ such that
$\var(e_i)=\var(c_{i \bs i})=1$, and we denote $\cov(e_i,c_{i \bs i})$ by $\ddot{\rho}_i$. We define 
$E_i^\nat = F_{E_i}^{-1} \circ \varphi(e_i)$
and $C_{i \bs i}^\nat=F_{C_{i \bs i}}^{-1} \circ \varphi(c_{i \bs i})$. The value of $\ddot{\rho}_i$ is then determined 
in one of 
four alternative ways: $\ddot{\rho}_i = \rho_i$; or so that the Pearson, Spearman or Kendall correlation of 
$E_i^\nat$ and $C_{i \bs i}^\nat$ matches that of $E_i$ and $C_{i \bs i}$. The choice of bivariate normal 
$(e_i,c_{i \bs i})$ ensures that $E_i^\nat$ and $C_{i \bs i}^\nat$ are symmetrically tail associated in the sense
described in the Introduction.
%***DAN: This means you have to introduce this well, and explain why the way we define tail assoc is
%the *right* way
For those familar with the language of copula statistics (see, e.g., \citealp{Ghosh_2020_AER} and \citealp{Genest_2007}), 
we replaced the copula of $(E_i,C_{i \bs i})$ with a normal copula to form $(E_i^\nat,C_{i \bs i}^\nat)$.
The choice of $\ddot{\rho}_i$ ensures that $E_i^\nat$ and $C_{i \bs i}^\nat$ are correlated to the same extent as 
$E_i$ and $C_{i \bs i}$, according to whichever definition of correlation is considered preferable. Thus $E_i^\nat$ and 
$C_{i \bs i}^\nat$ have had their tail association rendered symmetric while retaining their correlation \emph{per se},
as desired. The random variable $(E_j^\nat, C_{j \bs i}^\nat)$ is created in an analogous way.

@@

%G3
To acheive goal G3 from the Introduction i.e., to explore, using the lottery model, the likelihood that species can coexist with
ATAs but not without, or vice versa, we plotted the IGR, $\olr_{i \bs i}=\olr_{i \bs i} -q_{ij}\olr_{j \bs i}$, 
as well as the
IGR minus its ATA contribution, $\olr_{i \bs i} -\Delta_i^{[E \nat C]}$. 
For fixed values of $\sigma$ and $\delta$ and for left-tail associated noise, 
we were interested in the values of $\mu_1-\mu_2$ for which one of these
quantities is positive and the other is negative; this set of value characterizes the species which can coexist
with ATAs, but not without, or vice versa. The one dimensional space of values $\mu$ is a species trait space,
and $\mu_1$ and $\mu_2$ can be regarded as two points in it. But we instead define the ``trait-difference space'' of values $\mu_1-\mu_2$. 
Points of the trait-difference space characterize possible combinations of traits of the two competing species in our model. For fixed values of 
$\delta$ and $\sigma$, we defined the subset, $\psi_{\text{ATA}}(\delta,\sigma)$, of trait-difference space to be the set of 
values $\mu_1 - \mu_2$ for which the the IGR, $\olr_{i \bs i}$, is positive. This 
is the subspace for which the two species in the lottery model coexist in the presence of ATAs.
We also defined the subset, $\psi_{\lnot\text{ATA}}(\delta,\sigma)$, of trait-difference space to be the set of values
$\mu_1 - \mu_2$ for which the metric $\olr_{i \bs i} -\Delta_i^{[E \nat C]}$ is positive.
This is the subspace for which the two species in the lottery model coexist in the absence of 
ATAs. The symbol $\lnot$ denotes logical negation, so $\lnot \text{ATA}$ is intended to indicate the
absence of ATAs. We defined the ``ATA rescue space" associated with given values of $\delta$ and $\sigma$ to be the 
set difference $\psi_{\text{ATA}}(\delta,\sigma) \setminus \psi_{\lnot\text{ATA}}(\delta,\sigma)$, i.e., the
values $\mu_1 - \mu_2$ for which species coexist in the presence of ATAs, but not in their absence. 
Similarly, we define ``ATA exclusion space'' to be the set difference 
$\psi_{\lnot\text{ATA}}(\delta,\sigma) \setminus \psi_{\text{ATA}}(\delta,\sigma)$,
i.e., $\mu_1 - \mu_2$ values for which coexistence occurs in the absence, but not the presence, of ATAs.
On plots of $\olr_{i \bs i}$ and $\olr_{i \bs i} -\Delta_i^{[E \nat C]}$
against $\mu_1 - \mu_2$, we highlighted the ATA rescue and exclusion spaces. We plotted against
$\mu_1 - \mu_2$ for all combinations of $\delta=0.2, 0.4, 0.6, 0.8$ and $\sigma=2,4,5,6$.

>>>>

%Goal para
We previously demonstrated \citep{Ghosh_2020_AER} that ATAs exist and are common in nature, and we argued above 
that it is plausible both that climate change may alter environmental ATAs, and that ATAs may influence species 
coexistence. Thus we seek to explore the consequences of ATAs for competition, coexistence, and the maintenance of 
biodiversity, by means of the following specific goals. 
G1) To augment MCT to show formally how ATA plays a role in coexistence. 
MCT decomposes a measure of coexistence into components due to each of several mechanisms. 
We address how ATA contributes to these components. 
We here examine mechanisms of coexistence that apply in models exhibiting solely temporal variation; spatial variation can be considered in future work.
G2) To explore, using the simple lottery model, the relative magnitudes of the contributions to coexistence that come
from ATAs compared to other, known mechanisms of coexistence. This is to help illuminate whether ATA contributions to coexistence 
are likely to be negligible or meaningfully large. 
G3) To explore, using our simple model, whether species can coexist with
ATAs but not without, or vice versa. This is to begin to illuminate whether ATA contributions to biodiversity maintenance 
may be negligible or meaningfully large.
G4) To provide an empirical example of species coexistence which highlights the role ATAs can play.
In the Discussion, we revisit the possibility that climate change will alter
ATAs of environmental variables and thereby influence competition and species coexistence.
Overall, our study presents a new mechanism of species coexistence which may be influenced by climate change, 
and a means of understanding its theoretical importance and its importance for real ecosystems.

@@

We here pursue goal G1 from the Introduction: to augment MCT to illuminate how ATA
plays a role in coexistence. For pedagogical clarity, we develop each step of our new theory for a general 
two-species model, and for the lottery model (Introduction).
MCT and its recent computational extensions \citep{Ellner_2016,Ellner_2019} 
quantify the contributions of 
multiple mechanisms to coexistence. But we show below
that, for purely temporal variation, only one mechanism relates to tail associations: storage effects. 
So we begin by defining storage effects for a general two species model and 
for the lottery model. Our mathematical notation parallels
that of \cite{Ellner_2016,Ellner_2019}.

For a general model, it is assumed that the growth rate $r_i(t)=\ln(N_i(t+1)/N_i(t))$ 
of species $i$ can be written as an increasing function of
an environment-dependent factor $E_i(t)$ and as a decreasing function of a quantity representing
competitive pressure on species $i$, $C_i(t)$. As for the lottery model example (Introduction), we here assume, for simplicity, 
that the $E_i(t)$ are iid through time, and likewise for the
$C_i(t)$. Though this assumption can be circumvented, we make the assumption because it simplifies the
presentation of new ideas.

@@

We now specify how to decompose storage effects into
contributions due to ATAs and due
to $EC$ correlation \emph{per se}. 
Storage effects for species $i$ were computed by comparing
differences between invader and resident mean growth rates for the original model and for a 
modified model for which the components of $(E_i,C_{i \bs i})$ and those of $(E_j,C_{j \bs i})$
were rendered independent of each other [see (\ref{eq:storage_effects_gen})]. 
Because these components were rendered 
completely independent in the modified model, storage effects quantify the contribution of the 
whole $EC$ relationship, in all its aspects, to the invader/resident growth rate
difference. But we can instead consider distributions $(E_i^\nat,C_{i \bs i}^\nat)$ and 
$(E_j^\nat,C_{j \bs i}^\nat)$ with the same
marginal distributions, respectively, as $(E_i,C_{i \bs i})$ and $(E_j,C_{j \bs i})$,
and with some but not all aspects of the relationship 
between the components altered relative to the original distributions. Specifically, asymmetries of tail association are removed -- see below for how this is done. We then define 
$\olrn_{i \bs i}$ and $\olrn_{j \bs i}$ based on a modified model using  $(E_i^\nat,C_{i \bs i}^\nat)$
and $(E_j^\nat,C_{j \bs i}^\nat)$, and we consider the decomposition 
\begin{align}
\Delta_i^{(EC)} &= [(\olr_{i\bs i} - q_{ij} \olr_{j \bs i})-(\olrn_{i \bs i}-q_{ij} \olrn_{j \bs i})]+
[(\olrn_{i \bs i}-q_{ij} \olrn_{j \bs i})-(\olrs_{i \bs i}-q_{ij} \olrs_{j \bs i})] \label{eq:storage_effects_further1}\\
&= \Delta_i^{[E \nat C]} + \Delta_i^{[EC]}.\label{eq:storage_effects_further}
\end{align}
The term $\Delta_i^{[E \nat C]}$ is the component of storage effects due to 
ATAs between $E_i$ and $C_{i \bs i}$, and between $E_j$ and $C_{j \bs i}$.
The notation $\nat$ was chosen because this new symbol is part of the symbol $\shp$ (i.e., two of its four lines), 
corresponding to the removal of part of the relationship between $E$ and $C$ components (namely, ATAs).
The new symbol should correspondingly be pronounced ``partial sharp''.
The term $\Delta_i^{[EC]}$ is the component of storage effects due to $EC$ correlation \emph{per se}, remaining after 
having removed the effects of ATAs.
The brackets in the superscripts of $\Delta_i^{[E \nat C]}$ and $\Delta_i^{[EC]}$ are intended to indicate that 
(\ref{eq:storage_effects_further}) is a decomposition within a decomposition, i.e., (\ref{eq:storage_effects_further}) is a decomposition
of storage effects, which are themselves a term within the decomposition provided by MCT \citep{Ellner_2019} -- see 
below for additional details.

How are asymmetries of tail association removed while leaving $E$ and $C$ marginal distributions and overall 
correlation unchanged? We assume, for simplicity, that the cumulative distribution functions (CDFs) of $E_i$ and 
$C_{i \bs i}$, which we denote $F_{E_i}$ and $F_{C_{i \bs i}}$, respectively, are strictly monotonic and continuous. 
Letting $\varphi$ be the CDF of a standard normal distribution, we consider the random variables
$\varphi^{-1} \circ F_{E_i} (E_i)$ and $\varphi^{-1} \circ F_{C_{i \bs i}}(C_{i \bs i})$, which are standard normally
distributed, although they are not necessarily jointly bivariate normal.  Let $\rho_i$ be the covariance of
these random variables. We then define a bivariate normal random variable $(e_i,c_{i \bs i})$ such that
$\var(e_i)=\var(c_{i \bs i})=1$, and we denote $\cov(e_i,c_{i \bs i})$ by $\ddot{\rho}_i$. We define 
$E_i^\nat = F_{E_i}^{-1} \circ \varphi(e_i)$
and $C_{i \bs i}^\nat=F_{C_{i \bs i}}^{-1} \circ \varphi(c_{i \bs i})$. The value of $\ddot{\rho}_i$ is then determined 
in one of 
four alternative ways: $\ddot{\rho}_i = \rho_i$; or so that the Pearson, Spearman or Kendall correlation of 
$E_i^\nat$ and $C_{i \bs i}^\nat$ matches that of $E_i$ and $C_{i \bs i}$. The choice of bivariate normal 
$(e_i,c_{i \bs i})$ ensures that $E_i^\nat$ and $C_{i \bs i}^\nat$ are symmetrically tail associated in the sense
described in the Introduction.
%***DAN: This means you have to introduce this well, and explain why the way we define tail assoc is
%the *right* way
For those familiar with the language of copula statistics (see, e.g., \citealp{Ghosh_2020_AER} and \citealp{Genest_2007}), 
we replaced the copula of $(E_i,C_{i \bs i})$ with a normal copula to form $(E_i^\nat,C_{i \bs i}^\nat)$.
The choice of $\ddot{\rho}_i$ ensures that $E_i^\nat$ and $C_{i \bs i}^\nat$ are correlated to the same extent as 
$E_i$ and $C_{i \bs i}$, according to whichever definition of correlation is considered preferable. Thus $E_i^\nat$ and 
$C_{i \bs i}^\nat$ have had their tail association rendered symmetric while retaining their correlation \emph{per se},
as desired. The random variable $(E_j^\nat, C_{j \bs i}^\nat)$ is created in an analogous way.

@@

%G3
To achieve goal G3 from the Introduction i.e., to explore, using the lottery model, the likelihood that species can coexist with
ATAs but not without, or vice versa, we plotted the IGR, $\olr_{i \bs i}=\olr_{i \bs i} -q_{ij}\olr_{j \bs i}$, 
as well as the
IGR minus its ATA contribution, $\olr_{i \bs i} -\Delta_i^{[E \nat C]}$. 
For fixed values of $\sigma$ and $\delta$ and for left-tail associated noise, 
we were interested in the values of $\mu_1-\mu_2$ for which one of these
quantities is positive and the other is negative; this set of values characterizes the species which can coexist
with ATAs, but not without, or vice versa. The one dimensional space of values $\mu$ is a species trait space,
and $\mu_1$ and $\mu_2$ can be regarded as two points in it. But we instead define the ``trait-difference space'' of values $\mu_1-\mu_2$. 
Points of the trait-difference space characterize possible combinations of traits of the two competing species in our model. For fixed values of 
$\delta$ and $\sigma$, we defined the subset, $\psi_{\text{ATA}}(\delta,\sigma)$, of trait-difference space to be the set of 
values $\mu_1 - \mu_2$ for which the the IGR, $\olr_{i \bs i}$, is positive. This 
is the subspace for which the two species in the lottery model coexist in the presence of ATAs.
We also defined the subset, $\psi_{\lnot\text{ATA}}(\delta,\sigma)$, of trait-difference space to be the set of values
$\mu_1 - \mu_2$ for which the metric $\olr_{i \bs i} -\Delta_i^{[E \nat C]}$ is positive.
This is the subspace for which the two species in the lottery model coexist in the absence of 
ATAs. The symbol $\lnot$ denotes logical negation, so $\lnot \text{ATA}$ is intended to indicate the
absence of ATAs. We defined the ``ATA rescue space" associated with given values of $\delta$ and $\sigma$ to be the 
set difference $\psi_{\text{ATA}}(\delta,\sigma) \setminus \psi_{\lnot\text{ATA}}(\delta,\sigma)$, i.e., the
values $\mu_1 - \mu_2$ for which species coexist in the presence of ATAs, but not in their absence. 
Similarly, we define ``ATA exclusion space'' to be the set difference 
$\psi_{\lnot\text{ATA}}(\delta,\sigma) \setminus \psi_{\text{ATA}}(\delta,\sigma)$,
i.e., $\mu_1 - \mu_2$ values for which coexistence occurs in the absence, but not the presence, of ATAs.
On plots of $\olr_{i \bs i}$ and $\olr_{i \bs i} -\Delta_i^{[E \nat C]}$
against $\mu_1 - \mu_2$, we highlighted the ATA rescue and exclusion spaces. We plotted against
$\mu_1 - \mu_2$ for all combinations of $\delta=0.2, 0.4, 0.6, 0.8$ and $\sigma=2,4,5,6$.
